---
title: "EPI3 - In Class Labs"
author: "JMC"
date: "Last Update: `r format(Sys.time(), '%m/%d/%y')`"
output:
  html_document: 
    toc: true
    toc_float: 
          collapsed: false
          smooth_scroll: false
---
## Lab 1 - Review & Surveys
```{r, warning=FALSE, message=FALSE}
#want to see the html page? go here: https://iamjmc.github.io/epid3_sph/
library(haven)
library(dplyr)
library(tidyr)
library(expss)
#using NYC CHS data - see here for codebook: https://www.nyc.gov/assets/doh/downloads/pdf/episrv/chs2019-codebook.pdf
chs19 <- read_sas("https://www1.nyc.gov/assets/doh/downloads/sas/episrv/chs2019_public.sas7bdat")
keepcols <- c("strata", "wt20_dual", "usborn", "everasthma", "birthsex", "agegroup", "smoker", "insured", "newrace6", "age45up", "age50up", "generalhealth", "education", "exercise19",  "wallmold", "colonoscopy10yr19")
mydata <- dplyr::select(chs19, keepcols)


mydata <-  mydata %>%
  mutate(eversmoker=case_when(
  mydata$smoker=="2" ~ 1,
  mydata$smoker=="3" ~ 1,
  mydata$smoker=="1" ~ 0
)) 

mydata <- transform(mydata,
newrace6 = recode_factor(newrace6, `1`="White/N Afri/MidEastern, non-Hispanic", 
                                                 `2`="Black, non-Hispanic", 
                                                 `3`="Hispanic", 
                                                 `4`="Asian/PI, non-Hispanic",
                                                 `5`="Other, non-Hispanic"),
            
agegroup = recode_factor(agegroup, `1`="18-24", 
                                   `2`="25-44", 
                                   `3`="45-64", 
                                   `4`="65+", 
                                   `.d`="Don't Know",
                                   `.r`= "Refused"),

education = recode_factor(education, `1`="Less than HS",
                                    `2`="High school grad",
                                    `3`="Some college",
                                    `4`="College graduate",
                                    `.d`="Don’t know",
                                    `.r`="Refused"),
eversmoker = recode_factor(eversmoker, `0`="Non Smoker",
                                   `1` ="Smoker"),
birthsex = recode_factor(birthsex, `1`="Male",
                                 `2`="Female"),
insured = recode_factor(insured, `1`="Yes, insured",
                                `2`="No, uninsured"),
everasthma= recode_factor(everasthma, `1` = "Yes, Asthma",
                                      `2`="No Asthma"),
usborn= recode_factor(usborn, `1` = "US Born",
                                      `2`="Foreign Born"),
exercise19= recode_factor(exercise19, `1` = "Yes",
                                      `2`="No"),
wallmold= recode_factor(wallmold, `1` = "Yes",
                                      `2`="No"),
colonoscopy10yr19 = recode_factor(colonoscopy10yr19, `1`="Less than 10 yrs ago",
                                  `2` = "Never or >=10yr")
)
mydata = apply_labels(mydata,
                      insured = "Insured?",
                      birthsex = "Birth Sex", 
                      colonoscopy10yr19 = "Colonoscopy in last 10 years"
)

#pesky missing asthma data - if I don't delete this then table 1 will not be able to be stratified
mydata <- mydata %>%
           filter(!is.na(everasthma))

#check content of the data
summary(mydata)

#two by two table with chi square
table(mydata$birthsex, mydata$eversmoker)
colSums(table(mydata$birthsex, mydata$eversmoker))
rowSums(table(mydata$birthsex, mydata$eversmoker))
prop.table(table(mydata$birthsex, mydata$eversmoker))
chisq.test(mydata$birthsex, mydata$eversmoker, correct=FALSE)

#3 way table
ftable(mydata$smoker, mydata$birthsex, mydata$insured)
#another way to do two by two
library(expss)
mydata %>% tab_cols(total(), birthsex) %>% 
      tab_cells(insured) %>% 
      tab_mis_val(3) %>% 
      tab_stat_cases() %>% 
      tab_last_sig_cases() %>% 
      tab_pivot()

#summary tools 2x2 table
library(summarytools)
library(magrittr)
mydata %$%  # Acts like with(tobacco, ...)
  ctable(x = birthsex, y = eversmoker,
         chisq = TRUE,
         OR    = TRUE,
         RR    = TRUE,
         headings = FALSE,
         prop = "c" ) %>%
  print(method = "render")
```

```{r, warning=FALSE, message=FALSE}
library(survey)
library(sjstats)
svy <-
  svydesign(
    ids = ~ 1,
    strat = ~ strata,
    weights =  ~ wt20_dual, 
    data = mydata, 
    nest = TRUE,
    na.rm = TRUE
  )
#summary(svy)
#see means within the survey
mean(mydata$smoker, na.rm=TRUE)
svymean(~eversmoker, design=svy, strata = ~strata, weight = ~wt20_dual, data = mydata, na.rm=TRUE)
#see smoker mean stratified by birthsex
svyby(~eversmoker, ~birthsex, svy, svymean, na=TRUE)
svytable(~eversmoker+birthsex, svy)
svychisq(~eversmoker+birthsex, svy)
```

## Lab 2 - Causality & DAGs
You are interested in assessing whether physical activity( exposure) is causally related to heart disease (outcome). 
1. What other variables would you want to include in your DAG? Create a DAG with these variables and identify variables needed for total and direct effects?
```{r}
library(dagitty)
g <- dagitty('
dag {
"wall mold" [pos="-0.433,-0.322"]
age [pos="0.846,0.050"]
asthma [outcome,pos="0.341,-0.153"]
education [pos="-0.435,0.122"]
exercise [pos="0.324,0.355"]
howlonginUS [pos="-1.692,-0.427"]
race [pos="-0.569,-0.519"]
secondhandsmoker [pos="-0.031,0.869"]
smoker [pos="-0.903,0.620"]
usborn [exposure,pos="-1.195,-0.148"]
"wall mold" -> asthma
age -> asthma
age -> exercise
asthma -> exercise
education -> asthma
education -> exercise
education -> secondhandsmoker
education -> usborn
howlonginUS -> "wall mold"
howlonginUS -> usborn
race -> "wall mold"
race -> asthma
race -> usborn
secondhandsmoker -> asthma
secondhandsmoker -> exercise
secondhandsmoker -> smoker
smoker -> asthma
smoker -> education
smoker -> usborn
usborn -> asthma
}
')
plot(g)
```

#Adjusting for total effects= Age, Gender, SES
#Adjusting for direct effects = Age, BMI, Diet, Gender, SES

## Lab 3- Hypothesis Testing, Model Development, and Presentation of Results
Just a note: borough would require a DUA and currently not available. The variables noisetraffic, and badairwarning do not exist in CHS19. I use Wall Mold 
```{r, warning=FALSE, message=FALSE}
#Draw a dag for association between USBORN and EVERASTHMA
l3dag <- dagitty('
dag {
age [pos="0.846,0.050"]
asthma [outcome,pos="0.341,-0.153"]
education [pos="-0.435,0.122"]
exercise [pos="0.324,0.355"]
howlonginUS [pos="-1.692,-0.427"]
race [pos="-0.569,-0.519"]
secondhandsmoker [pos="-0.031,0.869"]
smoker [pos="-0.903,0.620"]
usborn [exposure,pos="-1.195,-0.148"]
age -> asthma
age -> exercise
asthma -> exercise
education -> asthma
education -> exercise
education -> secondhandsmoker
education -> usborn
howlonginUS -> usborn
race -> asthma
race -> usborn
secondhandsmoker -> asthma
secondhandsmoker -> exercise
secondhandsmoker -> smoker
smoker -> asthma
smoker -> education
smoker -> usborn
usborn -> asthma
}
')
plot (l3dag)
```

adjusting for direct effect - education, howlonginUS, race, smoker OR education, race, smoker, wall mold
adjusting for total effect - education, howlonginUS, race, smoker OR education, race, smoker, wall mold
```{r,  warning=FALSE, message=FALSE}
#create table 1
library(table1)
label(mydata$everasthma) <-"Ever had asthma"
label(mydata$birthsex) <- "Birth Sex"
label(mydata$usborn) <-"US Born"
label(mydata$agegroup) <-"Age Group"
label(mydata$eversmoker) <-"Smoke cig"
label(mydata$exercise19) <-"Exercise in past 30 days"
label(mydata$wallmold) <-"Visible mold in last year"
label(mydata$newrace6) <- "Race"
label(mydata$education) <- "Education"
#function for pvalues
pvalue <- function(x, ...) {
    # Construct vectors of data y, and groups (strata) g
    y <- unlist(x)
    g <- factor(rep(1:length(x), times=sapply(x, length)))
    if (is.numeric(y)) {
        # For numeric variables, perform a standard 2-sample t-test
        p <- t.test(y ~ g)$p.value
    } else {
        # For categorical variables, perform a chi-squared test of independence
        p <- chisq.test(table(y, g))$p.value
    }
    # Format the p-value, using an HTML entity for the less-than sign.
    # The initial empty string places the output on the line below the variable label.
    c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}

table1(~usborn + newrace6 + agegroup + birthsex + eversmoker  + wallmold | everasthma, data=mydata, overall="Total", extra.col=list(`P-value`=pvalue))
```

```{r,  warning=FALSE, message=FALSE}
#create table 2
library(gtsummary)
library(broom)
mydata$everasthma <- relevel(mydata$everasthma, ref = 2) 
mydata$usborn <- relevel(mydata$usborn, ref = 2) #have to make sure i have correct reference group 
#unadjusted model - adjusting for total and direct effect - education, race, smoker
fit.usborn <- glm(everasthma~usborn, 
                  data = mydata, 
                  family="binomial")%>% 
tbl_regression(
  exponentiate = TRUE )

fit.education <-glm(everasthma~education,
                   data=mydata,
                   family="binomial") %>%
    tbl_regression(
    exponentiate = TRUE)

fit.race <-glm(everasthma~newrace6,
                   data=mydata,
                   family="binomial") %>%
    tbl_regression(
    exponentiate = TRUE)

fit.eversmoker <- glm(everasthma~eversmoker,
                   data=mydata,
                   family="binomial") %>%
    tbl_regression(
    exponentiate = TRUE)

fit.wallmold <- glm(everasthma~wallmold,
                   data=mydata,
                   family="binomial") %>%
    tbl_regression(
    exponentiate = TRUE)

#adjusted model
fit.adjusted <-glm(everasthma~usborn + education+newrace6+eversmoker,
                   data=mydata,
                   family="binomial") %>%
    tbl_regression(
    exponentiate = TRUE,
  )

unadjusted <- tbl_stack(list(fit.usborn, fit.education, fit.race, fit.eversmoker))

tbl_merge(list(unadjusted, fit.adjusted), tab_spanner  = c("Unadjusted Analysis", "Adjusted Analysis"))
```

## Lab 4 - Interaction
Exposure: Insurance EM: Education Outcome: Colonoscopy 

Knol & VanderWeele - Presenting EM 
1. Present RR, OR/RD with CI for each stratum of E and X with a single reference category
2. Present RR, OR/RD with CI for E within strata of X
3. Present measures of EM on both additive and multiplicative scales (with CI & P)
4. List confounders for which the relation of E and D was adjusted

Knol - Presenting Interaction
1. Present RR, OR/RD with CI for each stratum of E and B with single reference category 
2. Present RR, OR/RD for E on D within strata of B and of B on D within strata of E
3. Present measures of EM on both additive (RERI) and multiplicative scales 
4. List confounders for which relation between E/D was adjusted and for which relation between B/D were adjusted
```{r,  warning=FALSE, message=FALSE}
library(summarytools)
#make a variable for 0-45, 45+ 
mydata <-  mydata %>%
transform(mydata,
age50up = recode_factor(age50up, `1`="50-64 Yrs", 
                                      `2`="65+"
))
#1 Look at insurance status and colonoscopy in different strata of education
table1(~insured + colonoscopy10yr19+ education + age50up, data=mydata, overall="Total")
mydata %$%  # First examine insurance and education
  ctable(x=education, y=insured, 
         chisq = TRUE,
         OR    = TRUE,
         RR    = TRUE,
         headings = FALSE,
         useNA="no",
         prop = "c" ) %>%
  print(method = "render")

mydata %$%  # Next colonoscopy and age 45+ 
  ctable(x=age50up, y=colonoscopy10yr19, 
         chisq = TRUE,
         OR    = TRUE,
         RR    = TRUE,
         headings = FALSE,
         useNA="no", #hides missing values
         prop = "c" ) %>%
  print(method = "render")

#make tables by level of education
stby(data    = list(x = mydata$insured, y = mydata$colonoscopy10yr19), 
     INDICES = mydata$education, 
     FUN     = ctable,
     chisq=TRUE, #no OR/RR here because it is categorical
     useNA="no"
     )


#2 make new variable for college education
mydata <- mydata %>%
  mutate(college=case_when(
    mydata$education=="College graduate" ~ 1, 
    mydata$education!="College graduate" ~ 0
  )) 

mydata <- mydata %>% 
  mutate(hasinsurance=case_when(
    mydata$insured=="Yes, insured" ~ 1, 
    mydata$insured=="No, uninsured" ~ 0
  ))
mydata <- mydata %>%
  mutate(recentcolonoscopy=case_when(
    mydata$colonoscopy10yr19=="Less than 10 yrs ago" ~ 1, 
    mydata$colonoscopy10yr19=="Never or >=10yr" ~0
  )
  )


#make tables by college education
stby(data    = list(x = mydata$insured, y = mydata$colonoscopy10yr19), 
     INDICES = mydata$college, 
     FUN     = ctable,
     chisq=TRUE, 
     OR= TRUE,
     RR=TRUE,
     useNA="no"
     )
library(interactionR)
library(broom)
#stratifying education categories - surely there must be an easier way a la bysort education: log in STATA?? 
fit.strata <- glm(colonoscopy10yr19~ insured + strata(education), data=mydata, family=binomial(link="logit"))

fit.strata.LHS <- glm(colonoscopy10yr19~ insured , subset=education=="Less than HS", data=mydata, family=binomial(link="logit")) %>%
    tbl_regression(
    exponentiate = TRUE,
  )
fit.crude <- glm(colonoscopy10yr19~ insured, data=mydata, family=binomial(link="logit")) %>%
  tbl_regression(
  exponentiate = TRUE,
  )

fit.strata.HS <- glm(colonoscopy10yr19~ insured , subset=education=="High school grad", data=mydata, family=binomial(link="logit")) %>%
    tbl_regression(
    exponentiate = TRUE,
  )
fit.strata.SC <- glm(colonoscopy10yr19~ insured , subset=education=="Some college", data=mydata, family=binomial(link="logit")) %>%
    tbl_regression(
    exponentiate = TRUE,
  )
fit.strata.CG <- glm(colonoscopy10yr19~ insured , subset=education=="College graduate", data=mydata, family=binomial(link="logit")) %>%
    tbl_regression(
    exponentiate = TRUE,
  )
tbl_merge(list(fit.crude, fit.strata.LHS, fit.strata.HS, fit.strata.SC, fit.strata.CG), tab_spanner  = c("Crude OR", "Less than HS", "High School grad", "Some College", "College Grad"))
```

```{r}
#calculate risk differences between college
library(epiR)
library(riskCommunicator)
tab <- table(mydata$insured, mydata$colonoscopy10yr19, mydata$college)
epi.2by2(tab)

colon.education.risk <- gComp(data = mydata, Y= "colonoscopy10yr19", X= "insured", #you can add covariates here by Z=c()
                           subgroup="education",
                           outcome.type = "binary", R = 200)
summary(colon.education.risk)
#plotting! :) 
library(ggplot2)
ggplot(colon.education.risk$results.df %>% 
         filter(Parameter %in% c("Risk Difference", "Risk Ratio"))
) + 
  geom_pointrange(aes(x = Comparison, 
                      y = Estimate, 
                      ymin = `2.5% CL`, 
                      ymax = `97.5% CL`, 
                      color = Comparison), 
                  shape = 2
  ) + 
  coord_flip() + 
  facet_wrap(~Parameter, scale = "free") + 
  theme_bw() + 
  theme(legend.position = "none")

colon.college.risk <- gComp(data = mydata, Y= "colonoscopy10yr19", X= "insured", 
                           subgroup="college",
                           outcome.type = "binary", R = 200)
summary(colon.college.risk)
```

```{r, warning=FALSE}
library(interactionR)
#interaction 
#with interactionR you may run into a problem of "At least one of the exposure names you have identified cannot be found in your model - You have to change the variables to straight 0/1 instead of the 1/2 found in CHS data 
fit.interaction <- glm(recentcolonoscopy~hasinsurance*college, 
                   data=mydata, 
                   family = binomial(link = "logit")) 

interactionR(fit.interaction,
exposure_names = c("college", "hasinsurance"),
ci.type = "delta", ci.level = 0.95,
em = F,  #T for EMM, F for interaction
recode = T
) %>%
interactionR_table(p.value=TRUE) %>%
             knit_print()
#RERI<0 implies negative additive interaction
#Multiplicative<1 implies negative multiplicative interaction
interactionR(fit.interaction,
exposure_names = c("college", "hasinsurance"),
ci.type = "delta", ci.level = 0.95,
em = T, 
recode = T
) %>%
interactionR_table(p.value=TRUE) %>%
             knit_print()


```

## Lab 5 - Mediation




## Lab 6 - Selection Bias & Models 