---
title: "Title: EPI3 - In Class Labs"
author: "JMC"
date: "Last Update: `r format(Sys.time(), '%m/%d/%y')`"
output:
  html_document: default
---
## Lab 1
```{r, warning=FALSE, message=FALSE}
#want to see the html page? go here: https://iamjmc.github.io/epid3_sph/
library(haven)
library(dplyr)
library(tidyr)
library(expss)
#using NYC CHS data - see here for codebook: https://www.nyc.gov/assets/doh/downloads/pdf/episrv/chs2019-codebook.pdf
chs19 <- read_sas("https://www1.nyc.gov/assets/doh/downloads/sas/episrv/chs2019_public.sas7bdat")
keepcols <- c("strata", "wt20_dual", "usborn", "everasthma", "birthsex", "agegroup", "smoker", "insured", "newrace6", "age45up", "generalhealth", "education", "exercise19",  "wallmold", "colonoscopy10yr19")
mydata <- dplyr::select(chs19, keepcols)

mydata <-  mydata %>%
  mutate(eversmoker=case_when(
  mydata$smoker=="2" ~ 1,
  mydata$smoker=="3" ~ 1,
  mydata$smoker=="1" ~ 0
)) 

mydata <- transform(mydata,
newrace6 = recode_factor(newrace6, `1`="White/N Afri/MidEastern, non-Hispanic", 
                                                 `2`="Black, non-Hispanic", 
                                                 `3`="Hispanic", 
                                                 `4`="Asian/PI, non-Hispanic",
                                                 `5`="Other, non-Hispanic"),
            
agegroup = recode_factor(agegroup, `1`="18-24", 
                                   `2`="25-44", 
                                   `3`="45-64", 
                                   `4`="65+", 
                                   `.d`="Don't Know",
                                   `.r`= "Refused"),
education = recode_factor(agegroup, `1`="Less than HS",
                                    `2`="High school grad",
                                    `3`="Some college",
                                    `4`="College graduate",
                                    `.d`="Don’t know",
                                    `.r`="Refused"),
eversmoker = recode_factor(eversmoker, `0`="Non Smoker",
                                   `1` ="Smoker"),
birthsex = recode_factor(birthsex, `1`="Male",
                                 `2`="Female"),
insured = recode_factor(insured, `1`="Yes, insured",
                                `2`="No, uninsured"),
everasthma= recode_factor(everasthma, `1` = "Yes, Asthma",
                                      `2`="No Asthma"),
usborn= recode_factor(usborn, `1` = "US Born",
                                      `2`="Foreign Born"),
exercise19= recode_factor(exercise19, `1` = "Yes",
                                      `2`="No"),
wallmold= recode_factor(wallmold, `1` = "Yes",
                                      `2`="No"),
colonoscopy10yr19 = recode_factor(colonoscopy10yr19, `1`="Less than 10 yrs ago",
                                  `2` = " Never or >=10yr")
)
mydata = apply_labels(mydata,
                      insured = "Insured?",
                      birthsex = "Birth Sex"
)

#pesky missing asthma data
mydata <- mydata %>%
           filter(!is.na(everasthma))

#check content of the data
summary(mydata)

#two by two table with chi square
table(mydata$birthsex, mydata$eversmoker)
colSums(table(mydata$birthsex, mydata$eversmoker))
rowSums(table(mydata$birthsex, mydata$eversmoker))
prop.table(table(mydata$birthsex, mydata$eversmoker))
chisq.test(mydata$birthsex, mydata$eversmoker, correct=FALSE)

#3 way table
ftable(mydata$smoker, mydata$birthsex, mydata$insured)
#another way to do two by two
library(expss)
mydata %>% tab_cols(total(), birthsex) %>% 
      tab_cells(insured) %>% 
      tab_mis_val(3) %>% 
      tab_stat_cases() %>% 
      tab_last_sig_cases() %>% 
      tab_pivot()

#summary tools 2x2 table
library(summarytools)
library(magrittr)
mydata %$%  # Acts like with(tobacco, ...)
  ctable(x = birthsex, y = eversmoker,
         chisq = TRUE,
         OR    = TRUE,
         RR    = TRUE,
         headings = FALSE,
         prop = "c" ) %>%
  print(method = "render")
```

```{r, warning=FALSE, message=FALSE}
library(survey)
library(sjstats)
svy <-
  svydesign(
    ids = ~ 1,
    strat = ~ strata,
    weights =  ~ wt20_dual, 
    data = mydata, 
    nest = TRUE,
    na.rm = TRUE
  )
summary(svy)
#see means within the survey
mean(mydata$smoker, na.rm=TRUE)
svymean(~eversmoker, design=svy, strata = ~strata, weight = ~wt20_dual, data = mydata, na.rm=TRUE)
#see smoker mean stratified by birthsex
svyby(~eversmoker, ~birthsex, svy, svymean, na=TRUE)
svytable(~eversmoker+birthsex, svy)
svychisq(~eversmoker+birthsex, svy)
```

## Lab 2
You are interested in assessing whether physical activity( exposure) is causally related to heart disease (outcome). 
1. What other variables would you want to include in your DAG? Create a DAG with these variables and identify variables needed for total and direct effects?
```{r}
library(dagitty)
g <- dagitty('
dag {
"wall mold" [pos="-0.433,-0.322"]
age [pos="0.846,0.050"]
asthma [outcome,pos="0.341,-0.153"]
education [pos="-0.435,0.122"]
exercise [pos="0.324,0.355"]
howlonginUS [pos="-1.692,-0.427"]
race [pos="-0.569,-0.519"]
secondhandsmoker [pos="-0.031,0.869"]
smoker [pos="-0.903,0.620"]
usborn [exposure,pos="-1.195,-0.148"]
"wall mold" -> asthma
age -> asthma
age -> exercise
asthma -> exercise
education -> asthma
education -> exercise
education -> secondhandsmoker
education -> usborn
howlonginUS -> "wall mold"
howlonginUS -> usborn
race -> "wall mold"
race -> asthma
race -> usborn
secondhandsmoker -> asthma
secondhandsmoker -> exercise
secondhandsmoker -> smoker
smoker -> asthma
smoker -> education
smoker -> usborn
usborn -> asthma
}
')
plot(g)
```

#Adjusting for total effects= Age, Gender, SES
#Adjusting for direct effects = Age, BMI, Diet, Gender, SES

## Lab 3- Hypothesis Testing, Model Development, and Presentation of Results
Just a note: borough would require a DUA and currently not available. The variables noisetraffic, and badairwarning do not exist in CHS19. I use Wall Mold 
```{r, warning=FALSE, message=FALSE}
#Draw a dag for association between USBORN and EVERASTHMA
l3dag <- dagitty('
dag {
age [pos="0.846,0.050"]
asthma [outcome,pos="0.341,-0.153"]
education [pos="-0.435,0.122"]
exercise [pos="0.324,0.355"]
howlonginUS [pos="-1.692,-0.427"]
race [pos="-0.569,-0.519"]
secondhandsmoker [pos="-0.031,0.869"]
smoker [pos="-0.903,0.620"]
usborn [exposure,pos="-1.195,-0.148"]
age -> asthma
age -> exercise
asthma -> exercise
education -> asthma
education -> exercise
education -> secondhandsmoker
education -> usborn
howlonginUS -> usborn
race -> asthma
race -> usborn
secondhandsmoker -> asthma
secondhandsmoker -> exercise
secondhandsmoker -> smoker
smoker -> asthma
smoker -> education
smoker -> usborn
usborn -> asthma
}
')
plot (l3dag)
```

adjusting for direct effect - education, howlonginUS, race, smoker OR education, race, smoker, wall mold
adjusting for total effect - education, howlonginUS, race, smoker OR education, race, smoker, wall mold
```{r,  warning=FALSE, message=FALSE}
#create table 1
library(table1)
label(mydata$everasthma) <-"Ever had asthma"
label(mydata$birthsex) <- "Birth Sex"
label(mydata$usborn) <-"US Born"
label(mydata$agegroup) <-"Age Group"
label(mydata$eversmoker) <-"Smoke cig"
label(mydata$exercise19) <-"Exercise in past 30 days"
label(mydata$wallmold) <-"Visible mold in last year"
label(mydata$newrace6) <- "Race"
label(mydata$education) <- "Education"
#function for pvalues
pvalue <- function(x, ...) {
    # Construct vectors of data y, and groups (strata) g
    y <- unlist(x)
    g <- factor(rep(1:length(x), times=sapply(x, length)))
    if (is.numeric(y)) {
        # For numeric variables, perform a standard 2-sample t-test
        p <- t.test(y ~ g)$p.value
    } else {
        # For categorical variables, perform a chi-squared test of independence
        p <- chisq.test(table(y, g))$p.value
    }
    # Format the p-value, using an HTML entity for the less-than sign.
    # The initial empty string places the output on the line below the variable label.
    c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}

table1(~usborn + newrace6 + agegroup + birthsex + eversmoker  + wallmold | everasthma, data=mydata, overall="Total", extra.col=list(`P-value`=pvalue))
```

```{r,  warning=FALSE, message=FALSE}
#create table 2
library(gtsummary)
library(broom)
mydata$everasthma <- relevel(mydata$everasthma, ref = 2) 
mydata$usborn <- relevel(mydata$usborn, ref = 2) #have to make sure i have correct reference group 
#unadjusted model - adjusting for total and direct effect - education, race, smoker
fit.usborn <- glm(everasthma~usborn, 
                  data = mydata, 
                  family="binomial")%>% 
tbl_regression(
  exponentiate = TRUE )

fit.education <-glm(everasthma~education,
                   data=mydata,
                   family="binomial") %>%
    tbl_regression(
    exponentiate = TRUE)

fit.race <-glm(everasthma~newrace6,
                   data=mydata,
                   family="binomial") %>%
    tbl_regression(
    exponentiate = TRUE)

fit.eversmoker <- glm(everasthma~eversmoker,
                   data=mydata,
                   family="binomial") %>%
    tbl_regression(
    exponentiate = TRUE)

fit.wallmold <- glm(everasthma~wallmold,
                   data=mydata,
                   family="binomial") %>%
    tbl_regression(
    exponentiate = TRUE)

#adjusted model
fit.adjusted <-glm(everasthma~usborn + education+newrace6+smoker,
                   data=mydata,
                   family="binomial") %>%
    tbl_regression(
    exponentiate = TRUE,
  )

unadjusted <- tbl_stack(list(fit.usborn, fit.education, fit.race, fit.eversmoker))

tbl_merge(list(unadjusted, fit.adjusted), tab_spanner  = c("Unadjusted Analysis", "Adjusted Analysis"))
```

Lab 4 - Interaction
Exposure: Insurance
EM: Education
Outcome: Colonoscopy 
```{r,  warning=FALSE, message=FALSE}
label(mydata$colonoscopy10yr19) <- "Colonoscopy in last 10 years"
#make a variable for 0-45, 45+ 
mydata <-  mydata %>%
  mutate(age45plus=case_when(
  mydata$agegroup=="18-24"| mydata$agegroup=="25-44" ~ 0,
  mydata$agegroup=="45-64" | mydata$agegroup=="65+" ~ 1
)) 

mydata <- transform(mydata,
age45plus = recode_factor(age45plus, `0`="Under 45 plus", 
                                      `1`="Over 45 plus"
))

table(mydata$agegroup, mydata$age45plus)
table1(~insured + colonoscopy10yr19+ education + age45plus, data=mydata, overall="Total")
mydata %$%  # Acts like with(tobacco, ...)
  ctable(x=education, y=insured, z=age45plus,
         chisq = TRUE,
         OR    = TRUE,
         RR    = TRUE,
         headings = FALSE,
         prop = "c" ) %>%
  print(method = "render")
```